---
title: 'Report Exercise 3: Data Wrangling'
author: 'Noémie Wellinger'
date: "`r Sys.Date()`"
output: 
 html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
---

# Introduction 
Part of Open Science is that data is open access. However, data can come in all shaped and formats. For an easy analysis, data should be optimally tidy and in an open access format. Since this is often not the reality, this exercise aims to clean data and prepare it for a simply analysis in R.
The dataset used contains measurements of soil organic carbon measurements in experiments, where ecosystems are exposed to ambient (low) and elevated (high) CO2 concentrations. First, the dataset was tidied manually in Excel. Then, in R, the log response ratio (RR) for differently aggregated data, namely looking at differences between the experiments and the influence of time since the experiment started.

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE)
```

# Methods and Results
```{r libs, message=F, warning=F}
library(tidyverse)
library(lubridate)
library(knitr)
```

Read in the tidied dataset on soil organic carbon.
```{r read, include=FALSE}
soil_original <- readr::read_csv2("../data/database1.csv")
soil_original
```

First, the unnecessary variables are removed.
```{r reduce}
soil_reduced <- soil_original |>
  select(experiment, max_depth, time_yrs, starts_with("ambient"), starts_with("increased")) |>
  mutate(time_yrs = as.numeric(time_yrs))
```

Then, the log-response ratio RR for each measurement is calculated.
$RR = ln (\frac{increased CO^2}{ambient CO^2})$
```{r calc_RR}
soil_reduced <- soil_reduced |> 
  mutate(RR = log(increased_co2_mean/ambient_co2_mean))
```

The data is aggregated by experiment, the other variables are summarized by mean.
```{r experiment}
soil_per_experiment <- soil_reduced |>
  group_by(experiment) |> 
  summarise(ambient_co2_mean = mean(ambient_co2_mean), 
            increased_co2_mean = mean(increased_co2_mean),
            RR = mean(RR))

knitr::kable(soil_per_experiment, caption = " Table 1: Log-response ratio aggregated per experiment", digits = 4)
```

Aggregate data across all experiments for different years since the start of the experiment (variable *time_yrs*), distinguishing an early phase (<3 years since start), a mid-phase (3-6 years since start), and a late phase (>6 years since start). Three phase labels are added in a new variable `phase`.
```{r time}
soil_time <- 
  soil_reduced |>
  filter(time_yrs < 3) |> 
    mutate(phase = "early") |> #add phase label

bind_rows(
soil_reduced |>
  filter(time_yrs >= 3 & time_yrs <= 6) |> 
    mutate(phase = "mid")) |>

bind_rows(
soil_reduced |>   
  filter(time_yrs > 6) |> 
    mutate(phase = "late")
)
```

Aggregate per phase, summarising time_yrs, ambient CO^2 and increased CO^2 by mean, then calculate the response ration for each phase.
```{r phase}
soil_per_phase <- soil_time |>
  group_by(phase) |> 
  summarise(time_yrs = mean(time_yrs), 
            ambient_co2_mean = mean(ambient_co2_mean), 
            increased_co2_mean = mean(increased_co2_mean),
            RR = mean(RR, na.rm = T))

knitr::kable(soil_per_phase, caption = "Table 2: Log-response ratio aggregated per phase", digits = 4, "simple")
```

# Discussion

