---
title: 'Report Exercise 1: Cleaning data from elevated CO2 experiments'
author: 'Noémie Wellinger'
date: "`r Sys.Date()`"
output: html_document
---


The dataset used contains measurements of soil organic carbon measurements in experiments, where ecosystems are exposed to ambient (low) and elevated (high) CO2 concentrations. 
Beforehand, the dataset was tidied manually in Excel and is now analyzed in R. 
The goal is to calculate the the log response ratio (RR) for differently aggregated data, namely looking at differences between the experiments and the influence of time since the experiment started.


```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE)
```

```{r read, include=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
soil_original <- readr::read_csv2("../data/database1.csv")
soil_original
```


First, the unnecessary variables are removed.
```{r reduce}
soil_reduced <- soil_original |>
  select(experiment, max_depth, time_yrs, starts_with("ambient"), starts_with("increased")) |>
  mutate(time_yrs = as.numeric(time_yrs))
```
Then, the log-response ratio RR for each measurement is calculated.
$RR = ln (\frac{ambient CO^2}{increased CO^2})$

```{r calc_RR}
soil_reduced <- soil_reduced |> 
  mutate(RR = log(increased_co2_mean/ambient_co2_mean))
```

The data is aggregated by experiment, the other variables are summarized by mean.
```{r experiment}
soil_per_experiment <- soil_reduced |>
  group_by(experiment) |> 
  summarise(ambient_co2_mean = mean(ambient_co2_mean), 
            increased_co2_mean = mean(increased_co2_mean),
            RR = mean(RR))

knitr::kable(soil_per_experiment, caption = " Table 1: Log-response ratio aggregated per experiment", digits = 4)
```

Aggregate data across all experiments for different years since the start of the experiment (variable *time_yrs*), distinguishing an early phase (<3 years since start), a mid-phase (3-6 years since start), and a late phase (>6 years since start). Three phase labels are added in a new variable. *phase*
```{r time}
soil_time <- 
  soil_reduced |>
  filter(time_yrs < 3) |> 
    mutate(phase = "early") |> #add phase label

bind_rows(
soil_reduced |>
  filter(time_yrs >= 3 & time_yrs <= 6) |> 
    mutate(phase = "mid")) |>

bind_rows(
soil_reduced |>   
  filter(time_yrs > 6) |> 
    mutate(phase = "late")
)
 
soil_time
```
Aggregate per phase, summarising time_yrs, ambient CO^2 and increased CO^2 by mean, then calculate the response ration for each phase.
```{r phase}
soil_per_phase <- soil_time |>
  group_by(phase) |> 
  summarise(time_yrs = mean(time_yrs), 
            ambient_co2_mean = mean(ambient_co2_mean), 
            increased_co2_mean = mean(increased_co2_mean),
            RR = mean(RR, na.rm = T))

knitr::kable(soil_per_phase, caption = "Table 2: Log-response ratio aggregated per phase", digits = 4, "simple")
```

```{r}
soil_time
soil_time |>
     filter(phase == "early")
```

